Computable Compressed Matrices\\Suplementary Material

### 2013-09-13 manuscript.tex
atualizei

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Computable Compressed Matrices}
}
% Insert Author names, affiliations and corresponding author email.
\\
Crysttian Arantes Paix\~{a}o$^{1\ast}$, 
Fl\'{a}vio Code\c{c}o Coelho$^{2}$, 
\\
\bf{1} Crysttian Arantes Paix\~{a}o Applied Mathematics School, Getulio Vargas Foundation, Rio de Janeiro, RJ, Brazil
\\
\bf{2} Fl\'{a}vio Code\c{c}o Coelho Applied Mathematics School, Getulio Vargas Foundation, Rio de Janeiro, RJ, Brazil
\\
$\ast$ E-mail: Corresponding crysttian.paixao@fgv.br
\end{flushleft}

@@

\begin{figure}[ht]
  \centering
  \subfigure[$\alpha=1,\beta_1=1$]{
  \includegraphics[scale=0.25,angle=-90,clip]{fig01}
  }
  \subfigure[$\alpha=1,\beta=32$]{
  \includegraphics[scale=0.25,angle=-90,clip]{fig02}
  }
  \subfigure[$\alpha=32,\beta=1$]{
  \includegraphics[scale=0.25,angle=-90,clip]{fig03}
  }
  \subfigure[$\alpha=64,\beta=64$]{
  \includegraphics[scale=0.25,angle=-90]{fig04}
  }
  \caption{Histograms constructed from samples with 10,000 elements, generated from a Beta distribution. Below each histogram is possible to verify the parameters used.}
  \label{fig:01020304}
\end{figure}

@@

\begin{figure}[h]
  \centering
   \subfigure[SM-VLB  Difference ($D = \eta_1 - \eta_2$)]{ 
   \includegraphics[scale=0.18,clip]{fig07}
   }
   \subfigure[SM Efficiency  ($\eta_1$)]{
   \includegraphics[scale=0.18,clip]{fig06}
   }
   \subfigure[VBL Efficiency ($\eta_2$)]{
   \includegraphics[scale=0.18,clip]{fig05}
   }
   \subfigure[SM$\geq$VLB Efficiency ($\eta_1 \geq \eta_2$)]{
   \includegraphics[scale=0.18,clip]{fig08}
   }
  \caption{Comparing compression efficiency of methods 1 and 2. Color scale in (a), (b) and (c) represent average bit-length. In (a) we can see the difference $D = \eta_1 - \eta_2$. It can be seen that for most combination of $\alpha$ and $\beta$, $D<0$, meaning the second method is more efficent to compress a sample of numbers with bit-lengths coming from a $Beta(\alpha,\beta)$ distribution. However, there is a small region in parameter space, which is shown in white on (d), where the SM method is more efficient. This region corresponds to the dots in red in (a), where the average bit-length is higher. In panels (b) and (c), we can see the efficiencies of SM and VLB methods, respectively.}
  \label{fig:06070809}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \subfigure[$\alpha_1=1,\beta_1=1,\alpha_2=1,\beta_2=1$]{
  \includegraphics[scale=0.28,angle=-90,clip]{fig09}
  }
  \subfigure[$\alpha_1=1,\beta_1=32,\alpha_2=32,\beta_2=1$]{
  \includegraphics[scale=0.28,angle=-90,clip]{fig10}
  }
  \subfigure[$\alpha_1=32,\beta_1=32,\alpha_2=32,\beta_2=32$]{
  \includegraphics[scale=0.28,angle=-90,clip]{fig11}
  }
  \subfigure[$\alpha_1=64,\beta_1=32,\alpha_2=32,\beta_2=64$]{
  \includegraphics[scale=0.28,angle=-90]{fig12}
  }
  \caption{Histograms constructed from samples with 10,000 elements, generated from the mixture of two Beta distributions with $w=0.5$. Below each histogram are the parameters of the mixture.}
  \label{fig:09101112}
\end{figure}

\begin{figure}[h]
  \centering
  \subfigure[$\alpha_1=64,\beta_1=48,\alpha_2=1,\beta_2=48$]{
  \includegraphics[scale=0.28,angle=-90]{fig13}
  }
  \subfigure[$\alpha_1=16,\beta_1=46,\alpha_2=49,\beta_2=64$]{
  \includegraphics[scale=0.28,angle=-90]{fig14}
  }
  \subfigure[$\alpha_1=16,\beta_1=16,\alpha_2=16,\beta_2=49$]{
  \includegraphics[scale=0.28,angle=-90]{fig15}
  }
  \subfigure[$\alpha_1=1,\beta_1=16,\alpha_2=1,\beta_2=49$]{
  \includegraphics[scale=0.28,angle=-90]{fig16}
  }
  \caption{Histograms constructed from samples with 10,000 elements, generated from the mixture of two Beta distributions with $w=0.5$. Below each histogram are the parameters of the mixture.}
  \label{fig:13141516}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \subfigure[Efficiency histogram of the SM method ]{
  \includegraphics[scale=0.25,angle=-90]{fig17}
  \label{fig:17}
  }
  \subfigure[Efficiency histogram of the VLB method ]{
  \includegraphics[scale=0.25,angle=-90]{fig18}
  \label{fig:18}
  }
  \caption{Efficiency histograms of the SM and VLB methods. Note that the VLB method has a greater average efficiency than SM method .}
  \label{fig:1718}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \subfigure[Efficiencies of the SM and VLB methods, for constant bit-length matrices.]{\includegraphics[scale=0.3,clip]{fig19}\label{fig19}}
  \subfigure[$\eta_1-\eta_2$ for matrices of constant bit-length.]{\includegraphics[scale=0.3,clip]{fig20}\label{fig20}}
  \caption{Compression efficiency of the SM and VLB methods for matrices of constant bit-length.}
  \label{fig:1920}
\end{figure}

@@

\section{Conclusion}
In this paper, we have focused in the compression of matrix data, since this is one of the most important application the authors foresee. However, the compression methodology presented can be applied to any numerical data structure, with gains to performance and memory footprint[citar tese Crysttian e possíveis artigos derivados].

Further discussions about doing computation with such compressed data-structures will be the subject of another manuscript (in preparation) in which we will present details about the implementation of the compression algorithm, and benchmarks on classical linear algebra tasks such as those in Linpack[ref].

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We would like to thank Claudia Torres Code\c{c}o, Paulo Cezar Carvalho and Moacyr Silva for fruitful discussions and key ideas which helped improve the manuscript.

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Computable Bitstring-compressed Matrices}
}
% Insert Author names, affiliations and corresponding author email.
\\
Crysttian Arantes Paix\~{a}o$^{1\ast}$, 
Fl\'{a}vio Code\c{c}o Coelho$^{2}$, 
\\
\bf{1} Crysttian Arantes Paix\~{a}o Applied Mathematics School, Getulio Vargas Foundation, Rio de Janeiro, RJ, Brazil
\\
\bf{2} Fl\'{a}vio Code\c{c}o Coelho Applied Mathematics School, Getulio Vargas Foundation, Rio de Janeiro, RJ, Brazil
\\
$\ast$ E-mail: Corresponding crysttian.paixao@fgv.br
\end{flushleft}

@@

\begin{figure}[ht]
  \centering
  \subfigure[$\alpha=1,\beta_1=1$]{
%   \includegraphics[scale=0.25,angle=-90,clip]{fig01}
  }
  \subfigure[$\alpha=1,\beta=32$]{
%   \includegraphics[scale=0.25,angle=-90,clip]{fig02}
  }
  \subfigure[$\alpha=32,\beta=1$]{
%   \includegraphics[scale=0.25,angle=-90,clip]{fig03}
  }
  \subfigure[$\alpha=64,\beta=64$]{
%   \includegraphics[scale=0.25,angle=-90]{fig04}
  }
  \caption{Histograms constructed from samples with 10,000 elements, generated from a Beta distribution. Below each histogram is possible to verify the parameters used.}
  \label{fig:01020304}
\end{figure}

@@

\begin{figure}[h]
  \centering
   \subfigure[SM-VLB  Difference ($D = \eta_1 - \eta_2$)]{ 
%    \includegraphics[scale=0.18,clip]{fig07}
   }
   \subfigure[SM Efficiency  ($\eta_1$)]{
%    \includegraphics[scale=0.18,clip]{fig06}
   }
   \subfigure[VBL Efficiency ($\eta_2$)]{
%    \includegraphics[scale=0.18,clip]{fig05}
   }
   \subfigure[SM$\geq$VLB Efficiency ($\eta_1 \geq \eta_2$)]{
%    \includegraphics[scale=0.18,clip]{fig08}
   }
  \caption{Comparing compression efficiency of methods 1 and 2. Color scale in (a), (b) and (c) represent average bit-length. In (a) we can see the difference $D = \eta_1 - \eta_2$. It can be seen that for most combination of $\alpha$ and $\beta$, $D<0$, meaning the second method is more efficent to compress a sample of numbers with bit-lengths coming from a $Beta(\alpha,\beta)$ distribution. However, there is a small region in parameter space, which is shown in white on (d), where the SM method is more efficient. This region corresponds to the dots in red in (a), where the average bit-length is higher. In panels (b) and (c), we can see the efficiencies of SM and VLB methods, respectively.}
  \label{fig:06070809}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \subfigure[$\alpha_1=1,\beta_1=1,\alpha_2=1,\beta_2=1$]{
%   \includegraphics[scale=0.28,angle=-90,clip]{fig09}
  }
  \subfigure[$\alpha_1=1,\beta_1=32,\alpha_2=32,\beta_2=1$]{
%   \includegraphics[scale=0.28,angle=-90,clip]{fig10}
  }
  \subfigure[$\alpha_1=32,\beta_1=32,\alpha_2=32,\beta_2=32$]{
%   \includegraphics[scale=0.28,angle=-90,clip]{fig11}
  }
  \subfigure[$\alpha_1=64,\beta_1=32,\alpha_2=32,\beta_2=64$]{
%   \includegraphics[scale=0.28,angle=-90]{fig12}
  }
  \caption{Histograms constructed from samples with 10,000 elements, generated from the mixture of two Beta distributions with $w=0.5$. Below each histogram are the parameters of the mixture.}
  \label{fig:09101112}
\end{figure}

\begin{figure}[h]
  \centering
  \subfigure[$\alpha_1=64,\beta_1=48,\alpha_2=1,\beta_2=48$]{
%   \includegraphics[scale=0.28,angle=-90]{fig13}
  }
  \subfigure[$\alpha_1=16,\beta_1=46,\alpha_2=49,\beta_2=64$]{
%   \includegraphics[scale=0.28,angle=-90]{fig14}
  }
  \subfigure[$\alpha_1=16,\beta_1=16,\alpha_2=16,\beta_2=49$]{
%   \includegraphics[scale=0.28,angle=-90]{fig15}
  }
  \subfigure[$\alpha_1=1,\beta_1=16,\alpha_2=1,\beta_2=49$]{
%   \includegraphics[scale=0.28,angle=-90]{fig16}
  }
  \caption{Histograms constructed from samples with 10,000 elements, generated from the mixture of two Beta distributions with $w=0.5$. Below each histogram are the parameters of the mixture.}
  \label{fig:13141516}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \subfigure[Efficiency histogram of the SM method ]{
%   \includegraphics[scale=0.25,angle=-90]{fig17}
  \label{fig:17}
  }
  \subfigure[Efficiency histogram of the VLB method ]{
%   \includegraphics[scale=0.25,angle=-90]{fig18}
  \label{fig:18}
  }
  \caption{Efficiency histograms of the SM and VLB methods. Note that the VLB method has a greater average efficiency than SM method .}
  \label{fig:1718}
\end{figure}

@@

\begin{figure}[h]
  \centering
%   \subfigure[Efficiencies of the SM and VLB methods, for constant bit-length matrices.]{\includegraphics[scale=0.3,clip]{fig19}\label{fig19}}
%   \subfigure[$\eta_1-\eta_2$ for matrices of constant bit-length.]{\includegraphics[scale=0.3,clip]{fig20}\label{fig20}}
  \caption{Compression efficiency of the SM and VLB methods for matrices of constant bit-length.}
  \label{fig:1920}
\end{figure}

@@

\section{Conclusion}
In this paper, we have focused in the compression of matrix data, since this is
one of the most important application the authors foresee. However, the
compression methodology presented can be applied to any numerical data
structure, with gains to performance and memory footprint\cite{teseCrysttian}.

Further discussions about doing computation with such compressed data-structures
will be the subject of another manuscript (in preparation) in which we will
present details about the implementation of the compression algorithm, and
benchmarks on classical linear algebra tasks such as those in
Linpack\cite{dongarra1979linpack}.

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We would like to thank Claudia Torres Code\c{c}o, Moacyr Silva and Paulo
Cezar Carvalho for fruitful discussions and key ideas which helped improve the
manuscript.

### 2013-09-13 suppl.tex
adicionei referência à figura.
edtei título
atualizei

\title{Computable Compressed Matrices\\Suplementary Material}
\author{Crysttian A. Paixão}

@@

\begin{abstract}
Your abstract.
\end{abstract}

@@

\paragraph{Subtraction}
Depending of the number involved in this operation, the result may be negative.
This fact alone generates the need for an extra bit to represent the sign($+$ or
$-$). The usual way to achieve this is to use the  two's complement
\cite{flores1963logic} representation which has the advantage of making the
addition subtraction and multiplication operation the same as those for unsigned
binary numbers. To exemplify let's subtract $7$ from $10$. To make use of the
same algorithm of the sum we represent the operation as $10+(-7)$. First we
need to convert the operands to two's complement representation (C2):

@@

This operation would be possible without the use of C2 representation as 10 is
greater than 7 (see table \ref{tab:04} for this).

@@

\begin{table}[ht]
	\centering
    \caption{Binary Multiplication}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binary & Bit-length \\
    \hline      
    $n_2$ & 10	& & 1010 & 3 [VERIFICAR]\\
    $n_1$ & \underline{$\times 7$} & & \underline{111} & 1 \\
    & & & 1010  & \\
    & & & \underline{1010\ } & \\
    & & & $11110$ & \\
    & & & \underline{1010\ \ } & \\
    Total& 70 & & 1000110 & 7\\
    \hline
	\end{tabular}
    \label{tab:06}
\end{table}

@@

If a signed integer is involved in the multiplication, we need to use C2
representation. To illustrate this case let's calculate $10\times-7$. in C2,
this opration becomes $01010\times11001$. Table \ref{tab:07} contains the
details. The basic difference is that the sign bits (in bold), positioned to
the left, are not involved in the operation. Only in the end they are used to
determine the sign of the product: $0$ meaning positive and $1$ negative. In
the worst case the bit-length of the product is the sum of the bit-lengths of
the factors plus 1 due to the sign.

\begin{table}[ht]
	\centering
    \caption{Multiplication involving a signed integer.}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binary & Bit-length \\
    \hline      
    $n_2$ & 10	& & \textbf{0}1010 & 5 \\
    $\times -1n_1$ & \underline{$\times -7$} & & \underline{\textbf{1}1001} & 5
\\
    & & & 1010  & \\
    & & & \underline{0000\ } & \\
    & & & $01010$ & \\
    & & & \underline{0000\ \ } & \\
    & & & 001010 & \\
    & & & \underline{1010\ \ \ } & \\
    Total& -70 & & \textbf{1}1011010 & 8\\
    \hline
	\end{tabular}
    \label{tab:07}
\end{table}

@@

\begin{table}[ht]
	\centering
    \caption{Binary division as a series of subtractions.}
    \begin{tabular}{crrrrr}
    \hline
    	  & Decimal & & Binary & Decimal quotient & Binary quotient\\
    \hline  
    $n_4$ & 50 					& & 110010
&0& 000 \\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 40      			& & 101000
&1& 001
\\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 30      			& & 011110
&2& 010
\\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 20      			& & 010100
&3& 011
\\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 10      			& & 001010
&4& 100
\\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 0      				& & 000000
&5& 101 \\
    \hline
	\end{tabular}
    \label{tab:07}
\end{table}

@@

@@

The operations on single numbers (scalars) as exposed above are a simplification
of the
actual operations taking place on the bitstrings as we compute with compressed
matrices. Additional details will be provided below when we discuss the
operation with matrices. Note that we are restricting the example to operations
which generate integer results. Handling of float operations and compression
will be the subject of a subsequent paper.

\begin{figure}[h]
 \centering
 \includegraphics[width=12cm]{./bitlength.png}
 % simples.png: 4800x3600 pixel, 600dpi, 20.32x15.24 cm, bb=0 0 576 432
 \caption{bitlength of the results for each of the four operations. The axes
represent the values of the operands.}
 \label{fig:bitlength}
\end{figure}

@@

After bitstring compression by the SM method, they become $3 \times 1$. They
are show in decimal and binary from in \ref{eq:01} and
\ref{eq:02}. The elements in $A$ and $B$ are 4 and 5 bits long, respectively
(the ``supreme minimum''). the ellipses ("$\ldots$") in the binary matrix
correspond the extra zeros to the left. Since the strings are written into 64
bits integers, sometimes we end up with unused bits to the left.

@@

\begin{equation}
	S = \begin{bmatrix}
			549649\\ 
			274889\\ 
			6181644
		\end{bmatrix}
        =
         \begin{bmatrix}
			\ldots000010000110001100010001\\
			\ldots000001000011000111001001\\
			\ldots010111100101001100001100
		\end{bmatrix}
        =
        \begin{bmatrix}

\ldots\underbrace{000010}_{2}\underbrace{000110}_{6}\underbrace{001100}_{12}
\underbrace{010001}_{17}\\	
 \ldots\underbrace{001101}_{13}\underbrace{010001}_{17}\underbrace{010100}_{20}
\underbrace{011000}_{24}\\
 \ldots\underbrace{010111}_{23}\underbrace{100101}_{37}\underbrace{001100}_{12}
\underbrace{001100}_{12}
		\end{bmatrix}
\end{equation}

@@

\begin{equation}\label{eq:ac2}
	-A = \begin{bmatrix}
			4952\\ 
			52841\\ 
			14251
		\end{bmatrix} 
        =
        \begin{bmatrix}
   			\ldots111111111101111011111000\\ 
			\ldots110100110010111010110111\\ 
			\ldots111101111001110110110101
		\end{bmatrix}
        =
        \begin{bmatrix}
 \ldots\underbrace{111111}_{-1}\underbrace{111101}_{-3}\underbrace{111011}_{-5}
\underbrace{111000}_{-8}\\
\ldots\underbrace{110100}_{-12}\underbrace{110010}_{-14}\underbrace{111010}_{-6}
\underbrace{110111}_{-9}\\
\ldots\underbrace{111101}_{-3}\underbrace{111001}_{-7}\underbrace{110110}_{-10}
\underbrace{110101}_{-11}
		\end{bmatrix}
\end{equation}

@@

\begin{equation}
	D = \begin{bmatrix}
			129\\ 
			14111238\\ 
			4685366
		\end{bmatrix} 
        =
        \begin{bmatrix}
   			\ldots000000000000000010000001 \\
			\ldots110101110101001000000110 \\
			\ldots010001110111111000110110
		\end{bmatrix}
        =
        \begin{bmatrix}
 \ldots\underbrace{000000}_{0}\underbrace{000000}_{0}\underbrace{000010}_{2}
\underbrace{000001}_{1} \\ 			 
 \ldots\underbrace{110101}_{-11}\underbrace{110101}_{-11}\underbrace{001000}_{8}
\underbrace{000110}_{6} \\
 \ldots\underbrace{010001}_{17}\underbrace{110111}_{23}\underbrace{111000}_{-8}
\underbrace{110110}_{-10}
		\end{bmatrix}
\end{equation}

@@

\ldots\underbrace{0011}_{3}\underbrace{0111}_{7}\underbrace{1010}_{10}
\underbrace{1011}_{11}
		\end{bmatrix}
\end{equation}
\begin{equation}
        \times\times \\
        \begin{bmatrix}

\ldots\underbrace{00001}_{1}\underbrace{00011}_{3}\underbrace{00111}_{7}
\underbrace{01001}_{9}  &

\ldots\underbrace{00001}_{1}\underbrace{00011}_{3}\underbrace{01110}_{14}
\underbrace{01111}_{15} &

\ldots\underbrace{10100}_{20}\underbrace{11110}_{30}\underbrace{00010}_{2}
\underbrace{00001}_{1}
		\end{bmatrix}
\end{equation}

To calculate $P[1,1]$ which in uncompressed form is $1 \times 1 + 3 \times 3 + 5
\times 7 + 8 \times 9$, we first must extract the numbers from the bitstrings
and then proceed with the linear combination. The extraction will make use of
masks as described before. Let tempA and tempB store the values of blocks 1 and
3 of AA[1,1] and BB[1,1]. Now we need to determine the size of the blocks
(bit-length) containing each element of the resulting matrix. As demonstrated
before, in the worst case, the product will have a bit-length which is the sum
of the bit-lengths of the operands. For this example this length is $4+5=9$,
however, we also have three additions for each element,adding a total of 3
extra bits, thus we end up with a bit-length of 12 for each element of the
product. Having determined the bit-length of the product, we can now we can do
the actual calculations and store the results.

@@

\begin{equation}
	P = \begin{bmatrix}
			1426882688\\ 
			2970686121\\ 
			3239350573
		\end{bmatrix} 
        =
        \begin{bmatrix}
   			\ldots000001010101000011001000000010000000 \\
			\ldots000010110001000100010001001010101001\\
			\ldots000011000001000101001001000100101101
		\end{bmatrix}
\end{equation}
\begin{equation}
        =
        \begin{bmatrix}

\ldots\underbrace{000001010101}_{117}\underbrace{000011001000}_{200}\underbrace{
000010000000}_{128}\\

\ldots\underbrace{000010110001}_{177}\underbrace{000100010001}_{273}\underbrace{
001010101001}_{681}\\

\ldots\underbrace{000011000001}_{193}\underbrace{000101001001}_{329}\underbrace{
000100101101}_{301} 			 
		\end{bmatrix}
\end{equation}

>>>>

\title{Computable Compressed Matrices\\Suplementary Material}
\author{Crysttian A. Paixão \and Flávio Codeço Coelho}

@@

@@

\paragraph{Subtraction}
Depending of the number involved in this operation, the result may be negative.
This fact alone generates the need for an extra bit to represent the sign ($+$
or
$-$). The usual way to achieve this is to use the  two's complement
\cite{flores1963logic} representation which has the advantage of making the
addition subtraction and multiplication operation the same as those for unsigned
binary numbers. This algorithm is what is actually implemented in CPUs. To
exemplify let's subtract $7$ from $10$. To make use of the
same algorithm of the sum , we represent the operation as $10+(-7)$. First we
need to convert the operands to two's complement representation (C2):

@@

This operation would be possible without the use of C2 representation as 10 is
greater than 7, and no signed integer is involved (see table \ref{tab:04} for
this).

@@

\begin{table}[ht]
	\centering
    \caption{Binary Multiplication}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binary & Bit-length \\
    \hline      
    $n_2$ & 10	& & 1010 & 4\\
    $n_1$ & \underline{$\times 7$} & & \underline{111} & 3 \\
    & & & 1010  & \\
    & & & \underline{1010\ } & \\
    & & & $11110$ & \\
    & & & \underline{1010\ \ } & \\
    Total& 70 & & 1000110 & 7\\
    \hline
	\end{tabular}
    \label{tab:06}
\end{table}

@@

If a signed integer is involved in the multiplication, we need to use C2
representation. To illustrate this case let's calculate $10\times-7$. in C2,
this operation becomes $01010\times11001$. Table \ref{tab:07} contains the
details. The basic difference is that the sign bits (in bold), positioned to
the left, are not involved in the operation. Only in the end they are used to
determine the sign of the product: $0$ meaning positive and $1$ negative. In
the worst case the bit-length of the product is the sum of the bit-lengths of
the factors plus 1 due to the sign.

\begin{table}[ht]
	\centering
    \caption{Multiplication involving a signed integer.}
    \begin{tabular}{crrrc}
    \hline
    	  & Decimal & & Binary & Bit-length \\
    \hline      
    $n_2$ & 10	& & \textbf{0}1010 & 5 \\
    $\times -1n_1$ & \underline{$\times -7$} & & \underline{\textbf{1}1001} & 5
\\
    & & & 1010  & \\
    & & & + \underline{0000\ } & \\
    & & & $\,01010$ & \\
    & & & + \underline{0000\ \ } & \\
    & & & 001010 & \\
    & & & + \underline{1010\ \ \ } & \\
    Total& -70 & & \textbf{1}1011010 & 8\\
    \hline
	\end{tabular}
    \label{tab:07}
\end{table}

@@

\begin{table}[ht]
	\centering
    \caption{Binary division as a series of subtractions.}
    \begin{tabular}{crrrrr}
    \hline
    	  & Decimal & & Binary & Decimal quotient & Binary quotient\\
    \hline  
    $n_4$ & 50 					& & 110010
&0& 000 \\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 40      			& & 101000
&1& 001
\\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 30      			& & 011110
&2& 010
\\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 20      			& & 010100
&3& 011
\\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 10      			& & 001010
&4& 100
\\
    $n_2$ & \underline{-10} 	& & \underline{001010}  &\underline{+1}&
\underline{+001} \\
    Remainder & 0      				& & 000000
&5& 101 \\
    \hline
	\end{tabular}
    \label{tab:08}
\end{table}

@@

Figure \ref{fig:bitlength} shows the bit-lengths of results of the four
operations with integer operands up to 8 bits in length. The conclusions listed
above, are visually emphasized in the figures.

@@

The operations on single numbers (scalars) as exposed above are a simplification
of the
actual operations taking place on the bitstrings as we compute with compressed
matrices. Additional details will be provided below, when we discuss the
operation with matrices. Note that we are restricting the example to operations
which generate integer results. Handling of float operations and compression
will be the subject of a subsequent paper.

\begin{figure}[h]
 \centering
%  \includegraphics[width=12cm]{./bitlength.png}
 % simples.png: 4800x3600 pixel, 600dpi, 20.32x15.24 cm, bb=0 0 576 432
 \caption{Bit-length of the results for each of the four operations. The axes
represent the values of the operands.}
 \label{fig:bitlength}
\end{figure}

@@

After bitstring compression by the SM method, they become $3 \times 1$. They
are show in decimal and binary from in \ref{eq:01} and
\ref{eq:02}. The elements in $A$ and $B$ are 4 and 5 bits long, respectively
(the ``supreme minimum''). The ellipses (``$\ldots$'') in the binary matrix
correspond the extra zeros to the left. Since the strings are written into 64
bits integers, sometimes we end up with unused bits to the left.

@@

\begin{align}
	S &= \begin{bmatrix}
			549649\\ 
			274889\\ 
			6181644
		\end{bmatrix}
        =
         \begin{bmatrix}
			\ldots000010000110001100010001\\
			\ldots000001000011000111001001\\
			\ldots010111100101001100001100
		\end{bmatrix}
      \\ \nonumber
      &= \begin{bmatrix}
\ldots\underbrace{000010}_{2}\underbrace{000110}_{6}\underbrace{001100}_{12}
\underbrace{010001}_{17}\\	
 \ldots\underbrace{001101}_{13}\underbrace{010001}_{17}\underbrace{010100}_{20}
\underbrace{011000}_{24}\\
 \ldots\underbrace{010111}_{23}\underbrace{100101}_{37}\underbrace{001100}_{12}
\underbrace{001100}_{12}
		\end{bmatrix}
\end{align}

@@

\begin{align}\label{eq:ac2}
	-A &= \begin{bmatrix}
			4952\\ 
			52841\\ 
			14251
		\end{bmatrix} 
        =
        \begin{bmatrix}
   			\ldots111111111101111011111000\\ 
			\ldots110100110010111010110111\\ 
			\ldots111101111001110110110101
		\end{bmatrix}
        \\\nonumber
        &= \begin{bmatrix}
 \ldots\underbrace{111111}_{-1}\underbrace{111101}_{-3}\underbrace{111011}_{-5}
\underbrace{111000}_{-8}\\
\ldots\underbrace{110100}_{-12}\underbrace{110010}_{-14}\underbrace{111010}_{-6}
\underbrace{110111}_{-9}\\
\ldots\underbrace{111101}_{-3}\underbrace{111001}_{-7}\underbrace{110110}_{-10}
\underbrace{110101}_{-11}
		\end{bmatrix}
\end{align}

@@

\begin{align}
	D &= \begin{bmatrix}
			129\\ 
			14111238\\ 
			4685366
		\end{bmatrix} 
        =
        \begin{bmatrix}
   			\ldots000000000000000010000001 \\
			\ldots110101110101001000000110 \\
			\ldots010001110111111000110110
		\end{bmatrix}
      \\\nonumber
        &=  \begin{bmatrix}
 \ldots\underbrace{000000}_{0}\underbrace{000000}_{0}\underbrace{000010}_{2}
\underbrace{000001}_{1} \\ 			 
 \ldots\underbrace{110101}_{-11}\underbrace{110101}_{-11}\underbrace{001000}_{8}
\underbrace{000110}_{6} \\
 \ldots\underbrace{010001}_{17}\underbrace{110111}_{23}\underbrace{111000}_{-8}
\underbrace{110110}_{-10}
		\end{bmatrix}
\end{align}

@@

\ldots\underbrace{0011}_{3}\underbrace{0111}_{7}\underbrace{1010}_{10}
\underbrace{1011}_{11}
		\end{bmatrix}
\end{equation}
\begin{small}
\begin{equation}
        \times\times \\
        \begin{bmatrix}\mathsf{
\ldots\underbrace{00001}_{1}\underbrace{00011}_{3}\underbrace{00111}_{7}
\underbrace{01001}_{9}}  &\mathsf{
\ldots\underbrace{00001}_{1}\underbrace{00011}_{3}\underbrace{01110}_{14}
\underbrace{01111}_{15}} & \mathsf{
\ldots\underbrace{10100}_{20}\underbrace{11110}_{30}\underbrace{00010}_{2}
\underbrace{00001}_{1}}
        \end{bmatrix}\nonumber
\end{equation}
\end{small}
To calculate $P[1,1]$ which in uncompressed form is $1 \times 1 + 3 \times 3 + 5
\times 7 + 8 \times 9$, we first must extract the numbers from the bitstrings
and then proceed with the linear combination. The extraction will make use of
masks as described before. Let tempA and tempB store the values of blocks 1 and
3 of AA[1,1] and BB[1,1]. Now we need to determine the size of the blocks
(bit-length) containing each element of the resulting matrix. As demonstrated
before, in the worst case, the product will have a bit-length which is the sum
of the bit-lengths of the operands. For this example this length is $4+5=9$,
however, we also have three additions for each element,adding a total of 3
extra bits, thus we end up with a bit-length of 12 for each element of the
product. Having determined the bit-length of the product, we can now we can do
the actual calculations and store the results.

@@

\begin{align}
	P &= \begin{bmatrix}
			1426882688\\ 
			2970686121\\ 
			3239350573
		\end{bmatrix} 
        =
        \begin{bmatrix}
   			\ldots000001010101000011001000000010000000 \\
			\ldots000010110001000100010001001010101001\\
			\ldots000011000001000101001001000100101101
		\end{bmatrix}
      \\ \nonumber
        &=        \begin{bmatrix}
\ldots\underbrace{000001010101}_{117}\underbrace{000011001000}_{200}\underbrace{
000010000000}_{128}\\
\ldots\underbrace{000010110001}_{177}\underbrace{000100010001}_{273}\underbrace{
001010101001}_{681}\\
\ldots\underbrace{000011000001}_{193}\underbrace{000101001001}_{329}\underbrace{
000100101101}_{301} 			 
        \end{bmatrix}
\end{align}
